\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}

### 2024-03-06 misc/JSS article/jss-article-rnw/article.tex

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()
fmean(mtcars$mpg)
fmean(EuStockMarkets)
fmean(mtcars)
fmean(mtcars$mpg, w = mtcars$wt)
fmean(mtcars$mpg, g = mtcars$cyl)
fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")

@@

R> airquality |> fmutate(

@@

+    ozone_deg = Ozone / Temp,

@@

  Ozone Solar.R Wind Temp Month Day rad_day ozone_deg ozone_amed ozone_resid
1    41     190  7.4   67     5   1   0.191    0.6119       TRUE     -10.279
2    36     118  8.0   72     5   2   0.135    0.5000       TRUE     -15.279
3    12     149 12.6   74     5   3   0.168    0.1622      FALSE      -3.035

@@

R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt

@@

R> mtcars |>
+    fsubset(mpg > 11) |>
+    fgroup_by(cyl, vs, am) |>

@@

R> mtcars |>
+    fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
+    fgroup_by(cyl, vs, am) |>
+    fmean(wt) |> head(2)

@@

R> mtcars |>
+   fgroup_by(vs) |>

@@

R> mtcars |>
+      fgroup_by(cyl, vs, am) |>

@@

Indexed by:  iso3c [1] | year [4 (61)] 
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties:
\begin{Schunk}
\begin{Sinput}
R> LIFEEXi = wldi$LIFEEX
R> str(LIFEEXi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:13176] 32.4 33 33.5 34 34.5 ...
 - attr(*, "label")= chr "Life expectancy at birth, total (years)"
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ iso3c: Factor w/ 216 levels "ABW","AFG","AGO",..: 2 2 2 2 2 2 ..
  .. ..- attr(*, "label")= chr "Country Code"
  ..$ year : Ord.factor w/ 61 levels "1960"<"1961"<..: 1 2 3 4 5 6 7..
  .. ..- attr(*, "label")= chr "Year"
\end{Soutput}
\begin{Sinput}
R> c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
\end{Sinput}
\begin{Soutput}
[1] FALSE  TRUE
\end{Soutput}
\begin{Sinput}
R> G(LIFEEXi[c(1:5, 7:10)])
\end{Sinput}
\begin{Soutput}
[1]    NA 1.590 1.544 1.494 1.448    NA 1.366 1.362 1.365

@@

By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. Users can request the generation of a \code{.join} column (\code{column = TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.

@@

R> df1 <- data.frame(id1 = c(1, 1, 2, 3),
+                    id2 = c("a", "b", "b", "c"),
+                    name = c("John", "Jane", "Bob", "Carl"),
+                    age = c(35, 28, 42, 50))
R> df2 <- data.frame(id1 = c(1, 2, 3, 3),
+                    id2 = c("a", "b", "c", "e"),

@@

An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}). Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.

@@

A few further particularities are worth highlighting. First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}. It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}) and defaults to \code{"left"}, so the default behavior simply adds columns to \code{x}. By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved. Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys. } Additionally, by default (\code{multiple = FALSE}), only the first matches from \code{y} are joined to avoid silent cartesian duplication of records. In multi-match settings, this will be reflected by few records from \code{y} being used. \fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided.

@@

A final noteworthy feature is the handling of duplicate non-id columns in both tables.

@@

The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both. Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions and incurring a high cost in terms of syntax and memory. \newline

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
      how = "longer", na.rm = FALSE, factor = c("names", "labels"),
      check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
      sort = FALSE, transpose = FALSE)
\end{Code}
%
The demonstration below employs a generated dataset about fruits. We could equivalently think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+              type_name = rep(c("Apples", "Bananas"), each = 2),
+              id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id       r      h
1    A    Apples  1 1.42361 1.7860
2    A    Apples  2 1.30998 0.0831
3    B   Bananas  1 0.04423 1.6729
4    B   Bananas  2 0.08639 1.2877
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named \code{lab\_name}. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the \code{variable} and \code{value} columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label   value
1    A    Apples  1        r Fruit Radius 1.42361
2    A    Apples  2        r Fruit Radius 1.30998
3    B   Bananas  1        r Fruit Radius 0.04423
4    B   Bananas  2        r Fruit Radius 0.08639
5    A    Apples  1        h Fruit Height 1.78599
6    A    Apples  2        h Fruit Height 0.08310
7    B   Bananas  1        h Fruit Height 1.67294
8    B   Bananas  2        h Fruit Height 1.28770
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id   r_A     r_B    h_A   h_B
1  1 1.424 0.04423 1.7860 1.673
2  2 1.310 0.08639 0.0831 1.288
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A       B
1  1        r Fruit Radius 1.4236 0.04423
2  2        r Fruit Radius 1.3100 0.08639
3  1        h Fruit Height 1.7860 1.67294
4  2        h Fruit Height 0.0831 1.28770
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

R> nest_lm_coef <- dl |>
+    rapply2d(lm, formula = mpg ~ .) |>
+    rapply2d(summary, classes = "lm") |>
+    get_elem("coefficients")

@@

Country-variance can be examined using \code{varying(wldi, effect = "year")}. For non-indexed data, \fct{varying} also has a \code{g} argument. A related exercise is to decompose the variance of a panel series into a component due to variation between countries and one capturing variance within countries over time. Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section~\ref{sec:ts_ps}, this is easily demonstrated

@@

wlddev |>
  subset(year >= 1990 & is.finite(GINI)) |>

@@

wlddev |>
  index_by(iso3c, year) |>

@@

The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse}, together with namespace masking, is able to provide a fast any syntactically clean experience of \proglang{R} - within GNU \proglang{R} - without the need to even restart the session. Masking is completely interactive and reversible within the active session: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove any \pkg{collapse} function from the list of exported functions, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project}\footnote{Website: https://fastverse.github.io/fastverse/} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and reattaches it at the top of the search path, letting its namespace to take precedence over other packages.

@@

        R 9.653 9.837 10.127  9.850 9.896 21.768   100
    Rfast 1.307 1.789  1.914  1.835 1.893  6.005   100
 collapse 1.322 1.431  1.566  1.474 1.577  4.987   100

@@

          R 10.75 11.213 11.706 11.281 11.472 16.423   100
 data.table 11.45 18.526 19.976 19.278 20.465 57.433   100
   collapse  1.92  2.713  3.303  2.959  3.552  6.251   100

@@

 data.table 136.21 137.08 138.09 137.64 138.34 143.8   100
   collapse  65.79  75.85  78.94  79.23  81.74 116.9   100

@@

 data.table 8.438 11.882 24.01 13.756 18.897 62.78   100
   collapse 1.221  1.382  1.48  1.408  1.466  2.94   100

@@

 data.table 11.32 14.70 21.36  16.00 17.63 58.90   100
   collapse 11.52 14.33 21.07  15.66 16.89 61.89   100

@@

 data.table 74.29 115.48 118.73 118.87 122.64 135.1   100
   collapse 55.69  87.16  91.14  90.35  94.12 118.8   100

@@

      base_int 60.216 63.192 64.81 64.170 65.81 108.20   100
  collapse_int  8.624  9.192 10.44  9.384 10.35  18.11   100
     base_char 92.110 94.566 98.56 97.757 99.91 160.08   100
 collapse_char 20.464 23.445 25.37 23.744 24.71 113.24   100

@@

        base_int 26.877 27.027 28.455 27.132 30.216 31.18    10
    collapse_int  8.637  8.701  9.269  8.722  8.973 13.14    10
       base_char 86.277 86.299 90.211 87.652 92.643 98.55    10
 data.table_char 41.150 41.615 42.225 41.806 41.906 47.05    10
   collapse_char 32.333 32.548 33.099 32.620 32.819 37.39    10

>>>>

with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects. As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()
R> fmean(mtcars$mpg)
[1] 20.09
R> fmean(EuStockMarkets)
 DAX  SMI  CAC FTSE 
2531 3376 2228 3566 
R> fmean(mtcars[5:10])
   drat      wt    qsec      vs      am    gear 
 3.5966  3.2173 17.8488  0.4375  0.4062  3.6875 
R> fmean(mtcars$mpg, w = mtcars$wt)
[1] 18.55
R> fmean(mtcars$mpg, g = mtcars$cyl)
    4     6     8 
26.66 19.74 15.10 
R> fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
    4     6     8 
25.94 19.65 14.81 
R> fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
   drat    wt  qsec     vs     am  gear
4 4.031 2.415 19.38 0.9149 0.6498 4.047
6 3.569 3.152 18.12 0.6212 0.3788 3.821
8 3.206 4.133 16.89 0.0000 0.1204 3.241
R> fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill") |> head(20)
 [1] 19.74 19.74 26.66 19.74 15.10 19.74 15.10 26.66 26.66 19.74 19.74 15.10
[13] 15.10 15.10 15.10 15.10 15.10 26.66 26.66 26.66

@@

R> airquality |> fmutate(ozone_deg = Ozone / Temp,

@@

@@

  Ozone Solar.R Wind Temp Month Day ozone_deg rad_day ozone_amed ozone_resid
1    41     190  7.4   67     5   1    0.6119   0.191       TRUE     -10.279
2    36     118  8.0   72     5   2    0.5000   0.135       TRUE     -15.279
3    12     149 12.6   74     5   3    0.1622   0.168      FALSE      -3.035

@@

R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt;

@@

R> mtcars |> fsubset(mpg > 11) |> fgroup_by(cyl, vs, am) |>

@@

R> mtcars |> fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
+    fgroup_by(cyl, vs, am) |> fmean(wt) |> head(2)

@@

R> mtcars |> fgroup_by(vs) |>

@@

R> mtcars |> fgroup_by(cyl, vs, am) |>

@@

Indexed by:  iso3c [1] | year [4 (61)] 
\end{Soutput}
\end{Schunk}
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties:
\begin{Schunk}
\begin{Sinput}
R> LIFEEXi <- wldi$LIFEEX; str(LIFEEXi, width = 70, strict = "cut")
\end{Sinput}
\begin{Soutput}
 'indexed_series' num [1:13176] 32.4 33 33.5 34 34.5 ...
 - attr(*, "label")= chr "Life expectancy at birth, total (years)"
 - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
  ..$ iso3c: Factor w/ 216 levels "ABW","AFG","AGO",..: 2 2 2 2 2 2 ..
  .. ..- attr(*, "label")= chr "Country Code"
  ..$ year : Ord.factor w/ 61 levels "1960"<"1961"<..: 1 2 3 4 5 6 7..
  .. ..- attr(*, "label")= chr "Year"
\end{Soutput}
\begin{Sinput}
R> c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
\end{Sinput}
\begin{Soutput}
[1] FALSE  TRUE
\end{Soutput}
\begin{Sinput}
R> G(LIFEEXi[c(1:5, 7:10)])
\end{Sinput}
\begin{Soutput}
[1]    NA 1.590 1.544 1.494 1.448    NA 1.366 1.362 1.365

@@

By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.

@@

R> df1 <- data.frame(id1 = c(1, 1, 2, 3), id2 = c("a", "b", "b", "c"),
+    name = c("John", "Jane", "Bob", "Carl"), age = c(35, 28, 42, 50))
R> df2 <- data.frame(id1 = c(1, 2, 3, 3), id2 = c("a", "b", "c", "e"),

@@

An alternative to the join column is to request an attribute (\code{attr = "name"/TRUE}) that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}). Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.

@@

A few further particularities are worth highlighting. First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}. It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}) and defaults to \code{"left"}, so the default behavior simply adds columns to \code{x}. By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved. Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys. } Additionally, by default (\code{multiple = FALSE}), only the first matches from \code{y} are joined to avoid silent cartesian duplication of records. In multi-match settings, this will be reflected by few records from \code{y} being used. \fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided:

@@

A final noteworthy feature is the handling of duplicate non-id columns in both tables:

@@

The reshaping/pivoting functionality of both commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both. Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory. \newline

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, factor = c("names", "labels"),
  check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
  sort = FALSE, transpose = FALSE)
\end{Code}
%
The demonstration below employs a generated dataset about fruits. We could equivalently think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
%
\begin{Schunk}
\begin{Sinput}
R> data <- data.frame(type = rep(c("A", "B"), each = 2),
+    type_name = rep(c("Apples", "Bananas"), each = 2),
+    id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
R> print(data)
\end{Sinput}
\begin{Soutput}
  type type_name id      r     h
1    A    Apples  1 0.3916 1.845
2    A    Apples  2 1.0761 1.667
3    B   Bananas  1 0.4958 1.976
4    B   Bananas  2 1.2679 2.974
\end{Soutput}
\begin{Sinput}
R> vlabels(data)
\end{Sinput}
\begin{Soutput}
          type      type_name             id              r              h 
            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
\end{Soutput}
\end{Schunk}
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named \code{lab\_name}. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the \code{variable} and \code{value} columns, respectively.
%
\begin{Schunk}
\begin{Sinput}
R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
\end{Sinput}
\begin{Soutput}
  type type_name id variable        label  value
1    A    Apples  1        r Fruit Radius 0.3916
2    A    Apples  2        r Fruit Radius 1.0761
3    B   Bananas  1        r Fruit Radius 0.4958
4    B   Bananas  2        r Fruit Radius 1.2679
5    A    Apples  1        h Fruit Height 1.8454
6    A    Apples  2        h Fruit Height 1.6666
7    B   Bananas  1        h Fruit Height 1.9762
8    B   Bananas  2        h Fruit Height 2.9737
\end{Soutput}
\begin{Sinput}
R> vlabels(dl)
\end{Sinput}
\begin{Soutput}
      type  type_name         id   variable      label      value 
        NA         NA "Fruit Id"         NA         NA         NA 
\end{Soutput}
\end{Schunk}
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} Note how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
\begin{Schunk}
\begin{Sinput}
R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
\end{Sinput}
\begin{Soutput}
  id    r_A    r_B   h_A   h_B
1  1 0.3916 0.4958 1.845 1.976
2  2 1.0761 1.2679 1.667 2.974
\end{Soutput}
\begin{Sinput}
R> namlab(dw)
\end{Sinput}
\begin{Soutput}
  Variable                  Label
1       id               Fruit Id
2      r_A  Fruit Radius - Apples
3      r_B Fruit Radius - Bananas
4      h_A  Fruit Height - Apples
5      h_B Fruit Height - Bananas
\end{Soutput}
\end{Schunk}
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
%
\begin{Schunk}
\begin{Sinput}
R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
+               labels = list(from = "type_name", to = "label"), how = "r"))
\end{Sinput}
\begin{Soutput}
  id variable        label      A      B
1  1        r Fruit Radius 0.3916 0.4958
2  2        r Fruit Radius 1.0761 1.2679
3  1        h Fruit Height 1.8454 1.9762
4  2        h Fruit Height 1.6666 2.9737
\end{Soutput}
\begin{Sinput}
R> vlabels(dr)
\end{Sinput}
\begin{Soutput}
        id   variable      label          A          B 
"Fruit Id"         NA         NA   "Apples"  "Bananas" 
\end{Soutput}
\end{Schunk}
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

@@

R> nest_lm_coef <- dl |> rapply2d(lm, formula = mpg ~ .) |>
+    rapply2d(summary, classes = "lm") |> get_elem("coefficients")

@@

Country-variance can be examined using \code{varying(wldi, effect = "year")}. For non-indexed data, \fct{varying} also has a \code{g} argument. A related exercise is to decompose the variance of a panel series into a component due to variation between countries and one capturing variance within countries over time. Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section~\ref{sec:ts_ps}, this is easily demonstrated:

@@

wlddev |> subset(year >= 1990 & is.finite(GINI)) |>

@@

wlddev |> index_by(iso3c, year) |>

@@

The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse}, together with namespace masking, is able to provide a fast and syntactically clean experience of \proglang{R} - without the need to even restart the session. Masking is completely interactive and reversible within the active session: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove any \pkg{collapse} function from the list of exported functions, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project}\footnote{Website: https://fastverse.github.io/fastverse/} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and reattaches it at the top of the search path, letting its namespace to take precedence over other packages.

@@

        R 9.827 9.846 10.453  9.872 10.069 21.22   100
    Rfast 1.316 1.792  2.352  1.857  2.110 11.40   100
 collapse 1.324 1.449  1.965  1.520  1.772 11.31   100

@@

          R 10.921 11.244 11.810 11.315 11.673  16.293   100
 data.table  8.183 18.755 21.582 19.753 21.508 113.610   100
   collapse  1.913  2.819  3.393  3.051  3.445   7.233   100

@@

 data.table 135.72 137.46 139.66 138.75 140.71 158.89   100
   collapse  63.92  74.84  79.86  79.75  85.08  96.31   100

@@

 data.table 8.930 13.689 29.964 16.518 57.245 125.224   100
   collapse 1.281  1.419  1.553  1.485  1.585   2.295   100

@@

 data.table 11.65 15.19 22.12  16.39 18.31 62.51   100
   collapse 11.67 15.02 21.86  16.05 17.61 63.43   100

@@

 data.table 69.41 116.29 120.08 119.79 123.5 143.8   100
   collapse 57.53  88.93  92.62  91.98  96.6 125.0   100

@@

      base_int 59.641 62.592 64.06 63.488 64.871 111.96   100
  collapse_int  8.413  9.133 10.11  9.291  9.694  15.32   100
     base_char 91.712 93.905 96.68 95.830 97.707 141.16   100
 collapse_char 21.339 22.828 24.25 23.053 23.813  71.40   100

@@

        base_int 26.730 26.743 28.290  26.89 30.47 31.18    10
    collapse_int  8.584  8.609  9.103   8.66  8.69 13.19    10
       base_char 86.599 87.263 90.344  88.16 91.71 98.66    10
 data.table_char 41.636 41.751 42.335  41.83 41.88 47.11    10
   collapse_char 32.563 32.793 33.322  32.88 32.94 37.40    10
